---
title: "Bayesian Futility Analysis for A/B Tests"
date: 2026-03-30
code-fold: false
echo: false
fig-cap-location: top
categories: []
number-sections: false
draft: true
cache: false
---

## Introduction

The sequential (or always valid) confidence interval we use at Datadog is great because experimenters can run an experiment until they detect an effect (albeit with a loss in power, but hey there is no free lunch).  But what if that day never comes?  

In a [previous post](../2026-02-20-time/index.qmd), I discussed how to think about experiment run times for Bayesian A/B tests.  The key result was a closed-form expression for the expected impact to the product as a function of total sample size, given a prior on the true lift and a decision rule of the form $S(N) < \hat \lambda$.  That post answered the question: _before seeing any data, how long should I plan to run?_

This post answers the complementary question: _given what I've seen so far, should I keep going?_ That is to say, this post concerns _futility_.  If at some interim point the probability of eventually reaching a ship decision is sufficiently low, we can stop the experiment early and save the traffic for something more promising.

Frequentists have a well-developed theory for this in the form of group sequential designs (I've written about group sequential designs [here](../2022-07-06-gsd/index.qmd) and [here](../2022-07-31-gsd-pt-2/index.qmd), although I didn't cover futility back then).  The Bayesian version is, in my opinion, more natural: we condition directly on the posterior and ask a straightforward probability question.


## Setup and Notation

Recall the setup from the previous post.  We target the lift

$$ \lambda = \dfrac{E[Y(1)]}{E[Y(0)]} - 1 \>. $$

We observe an estimate $\hat \lambda$ with standard error $s_\lambda$ (via the delta method), and use a conjugate normal-normal model with prior $\lambda \sim \operatorname{Normal}(\mu, \tau^2)$.  The posterior variance and mean are

$$ V^{-1} = \dfrac{1}{s^2_\lambda} + \dfrac{1}{\tau^2} \>, \qquad M = V \times \left(\dfrac{\hat \lambda}{s^2_\lambda} + \dfrac{\mu}{\tau^2}\right) \>. $$

The three decision rules from the previous post all could be expressed in the form

$$ S(N) < \hat \lambda \>, $$

where

$$ S(N) = s_\lambda^2 \left[ \Phi^{-1}(p) \sqrt{\frac{1}{\tau^2} + \frac{1}{s_\lambda^2}} - \frac{\mu}{\tau^2} \right] $$

and $p = c$ (probability to beat control), $p = 1-\alpha/2$ (credible interval), or $p = 0.5$ (posterior mean positive).

Suppose we have run the experiment for $K$ weeks, accruing $N_K$ total subjects.  We have observed $\hat \lambda_K$ with standard error $s_K$, giving us a posterior $(M_K, V_K)$.  If we continue for $W$ more weeks at a rate of $n_w$ subjects per week, the final sample size will be $N_f = N_K + W \cdot n_w$.  What is the probability we reach our shipping criteria after getting $W \cdot n_w$ more samples?


```{r helpers}
# =============================================================================
# Constants
# =============================================================================
sigma <- 0.5
prior_mean <- 0
prior_sd <- 0.05
control_mean <- 2.0
prob_treatment <- 0.5
N_K <- 10000

# =============================================================================
# Shared helpers (duplicated from the previous post for self-containment)
# =============================================================================

compute_se_lift <- function(n, prior_mean = 0, control_mean = 2.0,
                            prob_treatment = 0.5, sigma = 0.5) {
  se_c <- sigma / sqrt(n * (1 - prob_treatment))
  se_t <- sigma / sqrt(n * prob_treatment)
  treatment_mean <- control_mean * (1 + prior_mean)
  sqrt(se_t^2 / control_mean^2 + se_c^2 * treatment_mean^2 / control_mean^4)
}

closedform_impact_from_crit <- function(lambda_crit, se_lift,
                                         prior_mean = 0, prior_sd = 0.05) {
  tau <- sqrt(prior_sd^2 + se_lift^2)
  a   <- (prior_mean - lambda_crit) / tau
  prior_mean * pnorm(a) + (prior_sd^2 / tau) * dnorm(a)
}

# Compute posterior from observed summary statistics
compute_posterior <- function(lambda_hat, se_lambda, prior_mean = 0, prior_sd = 0.05) {
  V <- 1 / (1 / se_lambda^2 + 1 / prior_sd^2)
  M <- V * (lambda_hat / se_lambda^2 + prior_mean / prior_sd^2)
  list(M = M, V = V)
}

# Compute the ship threshold S(N) for a given decision rule
# p = 0.95 for P(lift > 0 | data) > 0.95
# p = 0.975 for 95% credible interval
# p = 0.5 for M > 0
compute_threshold <- function(n, p, prior_mean = 0, prior_sd = 0.05,
                               control_mean = 2.0, prob_treatment = 0.5, sigma = 0.5) {
  se_lift <- compute_se_lift(n, prior_mean, control_mean, prob_treatment, sigma)
  post_sd_inv <- sqrt(1 / prior_sd^2 + 1 / se_lift^2)
  se_lift^2 * (qnorm(p) * post_sd_inv - prior_mean / prior_sd^2)
}

prob_ship_if_continue <- function(M_K, V_K, N_f, p = 0.95,
                                   prior_mean = 0, prior_sd = 0.05,
                                   control_mean = 2.0, prob_treatment = 0.5,
                                   sigma = 0.5) {
  se_f <- compute_se_lift(N_f, prior_mean, control_mean, prob_treatment, sigma)
  S_f  <- compute_threshold(N_f, p, prior_mean, prior_sd,
                             control_mean, prob_treatment, sigma)
  pnorm((M_K - S_f) / sqrt(V_K + se_f^2))
}

expected_impact_if_continue <- function(M_K, V_K, N_f, p = 0.95,
                                         prior_mean = 0, prior_sd = 0.05,
                                         control_mean = 2.0, prob_treatment = 0.5,
                                         sigma = 0.5) {
  se_f <- compute_se_lift(N_f, prior_mean, control_mean, prob_treatment, sigma)
  S_f  <- compute_threshold(N_f, p, prior_mean, prior_sd,
                             control_mean, prob_treatment, sigma)
  closedform_impact_from_crit(S_f, se_f, prior_mean = M_K, prior_sd = sqrt(V_K))
}

sim_futility <- function(M_K, V_K, N_f, p = 0.95,
                          prior_mean = 0, prior_sd = 0.05,
                          control_mean = 2.0, prob_treatment = 0.5,
                          sigma = 0.5, B = 50000) {
  se_f <- compute_se_lift(N_f, prior_mean, control_mean, prob_treatment, sigma)
  S_f  <- compute_threshold(N_f, p, prior_mean, prior_sd,
                             control_mean, prob_treatment, sigma)

  true_lift    <- rnorm(B, M_K, sqrt(V_K))
  observed_lift <- rnorm(B, true_lift, se_f)

  mean(observed_lift > S_f)
}
```


## The Closed-Form Solution

We can avoid simulation entirely. The key insight is that the posterior $(M_K, V_K)$ plays exactly the same role as the prior $(\mu, \tau^2)$ did in the previous post.

Conditional on the current posterior, the true lift is $\lambda \sim \operatorname{Normal}(M_K, V_K)$ and the final observed lift given $\lambda$ is $\hat \lambda_f \mid \lambda \sim \operatorname{Normal}(\lambda, s_f^2)$.  Marginalizing over $\lambda$:

$$ \hat \lambda_f \sim \operatorname{Normal}(M_K, \> V_K + s_f^2) \>. $$

The ship decision requires $\hat \lambda_f > S(N_f)$, so

$$\Pr(\text{ship} \mid \text{continue until reaching } N_f) = \Phi\left(\frac{M_K - S(N_f)}{\sqrt{V_K + s_f^2}}\right) \>.$$

Seems pretty straight forward.  The nice part about this approach is that we can check how good our forecasts are since $\Pr(\text{ship} \mid \text{continue until reaching } N_f)$ can be empirically validated.  This brings us to one of my favorite topics to talk to data scientists about: calibration!


## Calibration, Such an Aggravation 

The closed-form $\Pr(\text{ship} \mid \text{continue until reaching } N_f)$ is a prediction. We should verify it is **calibrated**: among experiments where we predict $\Pr(\text{ship} \mid \text{continue until reaching } N_f) = p$, roughly fraction $p$ should actually ship.

To check this, we simulate the full lifecycle of many experiments end-to-end:

1. Draw a true lift from the prior.
2. Simulate interim data at week $K$ and compute the posterior $(M_K, V_K)$.
3. Compute the closed-form $\Pr(\text{ship} \mid \text{continue until reaching } N_f)$.
4. Simulate the actual remaining data and check whether the experiment actually ships.
5. Evaluate the calibration using something like `rms::val.prob` or similar tools.

You can see that below

```{r calibration}
#| fig-cap: "Calibration of the closed-form futility probability. We simulate 100,000 experiments with an interim analysis at $N_K = 1{,}000$ and a final analysis at $N_f = 5{,}000$. Each point represents a bin of experiments grouped by their predicted $\\Pr(\\text{ship} \\mid \\text{continue until reaching } N_f)$. The dashed line is perfect calibration."

set.seed(123)
nsims <- 100000
N_K <- 1000
N_f <- 5000
p_rule <- 0.95

se_K <- compute_se_lift(N_K)
se_f <- compute_se_lift(N_f)
S_f  <- compute_threshold(N_f, p = p_rule)

# 1. Draw true lifts from the prior
true_lift <- rnorm(nsims, prior_mean, prior_sd)

# 2. Simulate interim data
se_c_K <- sigma / sqrt(N_K * (1 - prob_treatment))
se_t_K <- sigma / sqrt(N_K * prob_treatment)
control_K   <- rnorm(nsims, control_mean, se_c_K)
treatment_K <- rnorm(nsims, control_mean * (1 + true_lift), se_t_K)
lambda_hat_K <- (treatment_K / control_K) - 1
se_lambda_K  <- sqrt(se_t_K^2 / control_K^2 + se_c_K^2 * treatment_K^2 / control_K^4)

# Compute interim posteriors
V_K_vec <- 1 / (1 / se_lambda_K^2 + 1 / prior_sd^2)
M_K_vec <- V_K_vec * (lambda_hat_K / se_lambda_K^2 + prior_mean / prior_sd^2)

# 3. Closed-form predicted P(ship)
predicted_p <- pnorm((M_K_vec - S_f) / sqrt(V_K_vec + se_f^2))

# 4. Simulate the actual final observed lift
# The final observed lift given the true lift has SE = se_f
# But we need the full final observed lift, not just the future part
# The final analysis sees all N_f subjects, so:
se_c_f <- sigma / sqrt(N_f * (1 - prob_treatment))
se_t_f <- sigma / sqrt(N_f * prob_treatment)
control_f   <- rnorm(nsims, control_mean, se_c_f)
treatment_f <- rnorm(nsims, control_mean * (1 + true_lift), se_t_f)
lambda_hat_f <- (treatment_f / control_f) - 1

# 5. Check if the experiment actually ships
actually_shipped <- lambda_hat_f > S_f
vp <- rms::val.prob(
  p = predicted_p, 
  y = actually_shipped
)



```

Now, you might say "_Demetri, you were simulating from the prior. I am not surprised that the procedure is calibrated_".  While correct, I think you are missing my point.  You can _check_ if your procedure is calibrated yourself!  You can even do this retrospectively.  Take a look at what you knew a week before an experiment ended.  Then, compute the probability you would have made a ship decision a week later, and _voila_.  This is a great way to check if your prior is "right" in my opinion (a topic I anticipate I will write a lot more on in the coming months).


## Bonus: Futility Boundary

We can also ask "At each interim week, what is the minimum posterior mean $M_K$ such that $\Pr(\text{ship} \mid \text{continue until reaching } N_f) > x\%$" for any given $x<0.5$?  This gives a **futility boundary** -- if your posterior mean is below the line at any point, stop the experiment!  Simply invert the win ship calculation to obtain

$$S(N_f) -  z_{1-x} \sqrt{V_k + s^2_f} +  \leq M_k \>. $$

```{r futility-boundary}
#| fig-cap: "Futility boundaries for three thresholds ($x = 0.1\\%$, $1\\%$, $10\\%$): the minimum posterior mean $M_K$ at each interim week such that $\\Pr(\\text{ship} \\mid \\text{continue until reaching } N_f) > x$ if the experiment continues to week 8. Decision rule: probability to beat control $> 0.95$."

max_weeks <- 8
n_per_week <- 5000
N_max <- max_weeks * n_per_week

interim_weeks <- 1:(max_weeks - 1)

thresholds <- list(
  list(x = 0.10,  label = "10%",  col = "firebrick"),
  list(x = 0.01,  label = "1%",   col = "darkorange"),
  list(x = 0.001, label = "0.1%", col = "steelblue")
)

se_f <- compute_se_lift(N_max)
S_f  <- compute_threshold(N_max, p = 0.95)

# Compute all boundaries to set ylim
all_boundaries <- lapply(thresholds, function(th) {
  sapply(interim_weeks, function(k) {
    N_K_local <- k * n_per_week
    se_K_local <- compute_se_lift(N_K_local)
    V_K_local <- 1 / (1 / se_K_local^2 + 1 / prior_sd^2)
    S_f + qnorm(th$x) * sqrt(V_K_local + se_f^2)
  })
})

ylim <- range(0, unlist(all_boundaries))

plot(NULL, xlim = range(interim_weeks), ylim = ylim,
     xlab = "Interim week", ylab = expression("Minimum" ~ M[K]))
grid()
abline(h = 0, lty = 2, col = 'gray50')

# Shade regions beneath each boundary (draw from most to least stringent
# so the narrower regions layer on top)
ymin <- ylim[1]
for (i in seq_along(thresholds)) {
  b <- all_boundaries[[i]]
  polygon(c(interim_weeks, rev(interim_weeks)),
          c(b, rep(ymin, length(interim_weeks))),
          col = adjustcolor(thresholds[[i]]$col, alpha.f = 0.2), border = NA)
}

for (i in seq_along(thresholds)) {
  lines(interim_weeks, all_boundaries[[i]], type = 'b', pch = 16,
        col = thresholds[[i]]$col, lwd = 2)
}

legend('bottomright',
       legend = sapply(thresholds, \(th) paste0("x = ", th$label)),
       col = sapply(thresholds, \(th) th$col),
       lwd = 2, pch = 16, fill = sapply(thresholds, \(th) adjustcolor(th$col, alpha.f = 0.2)),
       border = NA, cex = 0.8)
```


