---
title: Adaptive Shrinkage and the Student's t Prior
date: 2026-07-28
code-fold: false
echo: false
fig-cap-location: top
categories: []
number-sections: false
draft: false
cache: true
---


```{r}
library(tidyverse)
```

Consider the model

$$ \begin{align*} \hat \theta \mid \theta &\sim \operatorname{Normal}(\theta, s^2) \>, \\ \theta &\sim \operatorname{Normal}(\mu, \tau^2) \>.\end{align*} $$

The posterior mean for $\theta$ given $\hat \theta$ is 

$$
E[\theta \mid \hat\theta] =  \lambda \hat\theta + (1-\lambda)\mu,
\qquad \lambda = \frac{\tau^2}{s^2 + \tau^2} \>.
$$

In this form, it is straightforward to see that the posterior mean is a compromise between the prior mean and the data. Assuming the prior is fixed and that the prior mean is 0, $\mu = 0$ (as is often assumed in A/B testing), then the estimate $\hat \theta$ is shrunk toward 0 by a factor which depends only on the standard error and the prior standard deviation, and not at all on the point estimate.  If we let $r = s / \tau$ denote the ratio of uncertainty between the data and the prior, then the weight of the data is

$$ \lambda(r) = \dfrac{1}{1 + r^2} \>. $$

A value of $\lambda = 1$ means the data have all the weight, and $\lambda = 0$ means the prior has all the weight.  

```{r fig-normal-shrinkage}
#| fig-cap: |
#|   Weight of the data under a normal prior. The weight depends only on the ratio of
#|   uncertainty, never on the point estimate.

shrink_guide <- tibble(
  label   = c("More certain data", "Less certain data"),
  x       = c(0.85, 3.15),
  xend    = c(0.08, 3.92),
  label_x = c(0.95, 3.05),
  hjust   = c(0, 1),
  y       = 0.1
)

ggplot() +
  stat_function(fun = \(x) 1 / (1 + x^2), xlim = c(0, 4), n = 1000, linewidth = 1.2) +
  geom_segment(
    data = shrink_guide, aes(x = x, xend = xend, y = y, yend = y),
    arrow = arrow(length = unit(0.16, "inches"), type = "closed"),
    linewidth = 0.9, colour = "gray35"
  ) +
  # geom_label rather than geom_text so the curve, which crosses y = 0.1 near r = 3,
  # does not run through the text
  geom_label(
    data = shrink_guide, aes(x = label_x, y = y, label = label, hjust = hjust),
    vjust = 0.5, size = 3.3, colour = "gray25",
    fill = "white", linewidth = 0, label.padding = unit(0.12, "lines")
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.02))) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(
    x = expression("Ratio of uncertainty " * r == frac(s, tau)),
    y = expression("Weight of the data " * lambda(r))
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title = element_text(colour = "gray20")
  )
```

Shrinkage is good for a variety of reasons.  However, sometimes the shrinkage afforded by the conjugate normal can place the posterior in a region supported by neither the likelihood nor the prior -- a proverbial no-man's-land of parameter space.  Such cases are called *prior data conflict*, and happen when the data show a large effect which the prior deems unlikely.  @fig-prior-data-conflict shows an example.  The prior and likelihood are equally certain here, so $\lambda = 0.5$ and the posterior lands exactly halfway between them -- in a region where neither one places much density.  


```{r fig-prior-data-conflict}
#| fig-cap: |
#|   Prior data conflict. The posterior sits between the prior and the likelihood, in a
#|   region where neither places much density. Curves are scaled to a common height.

conflict <- tibble(
  component = c("Normal", "Likelihood"),
  mean      = c(0, 0.30),
  sd        = c(0.05, 0.05)
) |>
  # Conjugate posterior: precisions add, mean is the precision-weighted average
  add_row(
    component = "Posterior",
    mean = 0.30 * 0.05^2 / (0.05^2 + 0.05^2),
    sd = sqrt(1 / (1 / 0.05^2 + 1 / 0.05^2))
  )

conflict_curves <- conflict |>
  reframe(
    x = seq(-0.3, 0.8, length.out = 600),
    density = dnorm(x, mean, sd),
    .by = component
  ) |>
  # Scale each curve to a common peak height; the y axis is relative height, not density
  mutate(density = density / max(density), .by = component)

conflict_labels <- tibble(
  component = c("Normal", "Posterior", "Likelihood"),
  x         = c(-0.10, 0.15, 0.385),
  y         = c(1.09, 1.15, 1.09)
)

# Order the curves left to right so the sequential palette tracks position on the axis
component_levels <- c("Normal", "Posterior", "Likelihood")
conflict_curves <- mutate(
  conflict_curves, component = factor(component, component_levels)
)
conflict_labels <- mutate(
  conflict_labels, component = factor(component, component_levels)
)

ggplot(conflict_curves, aes(x, density, colour = component)) +
  geom_line(linewidth = 1.2) +
  geom_text(
    data = conflict_labels, aes(x = x, y = y, label = component),
    size = 3.8, show.legend = FALSE
  ) +
  scale_colour_viridis_d(end = 0.85) +
  scale_x_continuous(breaks = seq(-0.25, 0.75, 0.25)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  guides(colour = "none") +
  labs(
    x = expression("Effect size " * theta),
    y = "Relative density"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    # The curves are scaled to a common peak, so the y values carry no information
    axis.text.y = element_blank(),
    axis.title = element_text(colour = "gray20")
  )
```


Eventually the data will hold all the weight and this prior data conflict won't be such a big deal.  How fast does that happen?  The question is easiest to answer in odds rather than weights.  Rearranging $\lambda = 1/(1 + r^2)$,

$$ \dfrac{\lambda}{1 - \lambda} = \dfrac{1}{r^2} = \dfrac{\tau^2}{s^2} \>. $$

The left side is the odds that the posterior mean favors the data over the prior.  Since $s^2 \propto 1/n$, those odds are proportional to the effective sample size, so collecting $k$ times as much data multiplies them by exactly $k$.  Converting back to a weight,

$$ \lambda(kn) = \dfrac{k \, \lambda(n)}{1 + (k-1)\lambda(n)} \>, \qquad \dfrac{\lambda(kn)}{\lambda(n)} = \dfrac{k}{1 + (k-1)\lambda(n)} \>. $$

The consequence of increasing the sample can be read straight off the current weight of the data.

```{r fig-lambda-growth}
#| fig-cap: |
#|   Weight of the data after multiplying the effective sample size by k, for several
#|   starting values.

lambda_growth <- expand_grid(
  lambda_now = c(0.1, 0.2, 0.3, 0.5, 0.8),
  k = seq(1, 10, length.out = 400)
) |>
  mutate(lambda_new = k * lambda_now / (1 + (k - 1) * lambda_now))

growth_labels <- lambda_growth |>
  filter(k == max(k), .by = lambda_now)

ggplot(lambda_growth, aes(k, lambda_new, colour = factor(lambda_now))) +
  geom_line(linewidth = 1.2) +
  geom_text(
    data = growth_labels,
    aes(label = paste0("lambda(n) == ", lambda_now)),
    parse = TRUE, hjust = 0, nudge_x = 0.15, size = 3.6, show.legend = FALSE
  ) +
  scale_x_continuous(breaks = 1:10, expand = expansion(mult = c(0.01, 0.16))) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  scale_colour_viridis_d(end = 0.85, direction = -1) +
  guides(colour = "none") +
  labs(
    x = "Sample size multiplier k",
    y = expression("Weight of the data " * lambda(kn))
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.title = element_text(colour = "gray20"))
```

Let's focus on the curves above for a moment.  What they tell us is that "overwhelming the prior with the likelihood" can be _incredibly difficult_.  Take the $\lambda(n) = 0.5$ line, which corresponds to @fig-prior-data-conflict.  If I were to double my sample size, I would increase the weight of the data by only 17 percentage points.  That...doesn't feel like a good return on investment, especially if the current sample size is very large.  Moreover, you can see diminishing returns on getting more data.  Clearly, waiting for more data is not always a viable solution to prior data conflict.

As an aside, we can ask the question in reverse.  Because the odds scale with $n$, the sample size needed to reach a target $\lambda^\star$ is just the ratio of the two odds,

$$ \dfrac{N}{n} = \dfrac{\lambda^\star / (1 - \lambda^\star)}{\lambda(n) / (1 - \lambda(n))} \>. $$

Moving from $\lambda = 0.2$ to $\lambda = 0.9$ means going from odds of $0.25$ to odds of $9$: a thirty-six-fold increase in effective sample size.  In most experiments that is not a decision anyone gets to make.  If the conflict is uncomfortable at the current sample size, waiting for more data is not the fix -- the prior is.

## The Student's t prior

Consider an extension to the model above in which the prior for $\theta$ is now a Student's t distribution with degrees of freedom $\nu$, mean $\mu$, and standard deviation $\tau$:

$$ \begin{align*} \hat \theta \mid \theta &\sim \operatorname{Normal}(\theta, s^2) \>,\\ \theta &\sim \operatorname{Student-t}(\nu, \mu, \tau^2) \>.\end{align*} $$

This model lacks a closed form posterior, so the expectation is computed by quadrature.  Because the prior is heavier tailed, the weight placed on the data is now a function of the ratio of uncertainties, the degrees of freedom, _and_ the point estimate.  That last part is really important, and it means we get to plot the equivalent of @fig-normal-shrinkage in two dimensions.  I'll still plot $r$ on the $x$-axis, and to make things scale-invariant I will plot $z = \hat \theta / s$ on the $y$-axis.  The plots for a few degrees of freedom are shown below.

```{r}

lambda_breaks <- seq(0.1, 0.9, 0.2)

# Posterior mean for y ~ N(theta, 1) with a t_nu prior whose standard deviation is
# tau = 1 / r; the scale that delivers that standard deviation is sigma (needs nu > 2)
post_mean <- function(z, r, nu) {
  tau <- 1 / r
  sigma <- tau * sqrt((nu - 2) / nu)
  log_kern <- function(theta) {
    dt(theta / sigma, df = nu, log = TRUE) - log(sigma) + dnorm(z - theta, log = TRUE)
  }
  # The integrand has mass near 0 and near z; integrate each region separately so the
  # quadrature cannot step over a narrow spike
  brk <- sort(unique(c(0, z)))
  span <- 60 * max(sigma, 1)
  lo <- min(brk) - span
  hi <- max(brk) + span
  # Rescale by the peak: when the prior is tight and the estimate is far from zero, both
  # integrals underflow to 0 on the raw scale and the ratio comes back NaN
  probe <- c(outer(brk, c(0, sigma * c(-8, -2, 2, 8), c(-8, -2, 2, 8)), "+"), lo, hi)
  peak <- max(log_kern(probe))
  f <- function(theta) exp(log_kern(theta) - peak)
  piece <- function(g) {
    sum(mapply(
      \(a, b) integrate(g, a, b, subdivisions = 2000L, rel.tol = 1e-10,
                        stop.on.error = FALSE)$value,
      c(lo, brk), c(brk, hi)
    ))
  }
  piece(\(th) th * f(th)) / piece(f)
}

t_shrinkage <- expand_grid(
  nu = c(3, 10, 20, 50),
  r = seq(0.05, 4, length.out = 120),
  z = seq(0.05, 10, length.out = 120)
) |>
  mutate(lambda = pmap_dbl(list(z, r, nu), \(z, r, nu) post_mean(z, r, nu) / z))
```

```{r fig-t-shrinkage}
#| fig-cap: |
#|   Weight of the data under a mean-zero Student's t prior, with the prior standard
#|   deviation held at $\tau = s/r$ in every panel. Solid contours are the t prior at
#|   $\lambda$ = 0.1, 0.3, 0.5, 0.7, 0.9; dotted vertical lines are the normal prior at
#|   the same values.
#| fig-width: 9
#| fig-height: 7
ggplot(t_shrinkage, aes(r, z)) +
  geom_raster(aes(fill = lambda), interpolate = TRUE) +
  # The normal-prior contours are vertical because lambda = 1 / (1 + r^2) does not
  # involve the point estimate
  geom_vline(
    data = tibble(r = sqrt(1 / lambda_breaks - 1)),
    aes(xintercept = r), colour = "white", linewidth = 0.3, linetype = "22"
  ) +
  geom_contour(aes(z = lambda), colour = "white", linewidth = 0.5,
               breaks = lambda_breaks) +
  facet_wrap(~nu, labeller = label_bquote(nu == .(nu))) +
  scale_fill_viridis_c(limits = c(0, 1), name = expression(lambda)) +
  labs(x = expression("Ratio of uncertainty " * r == frac(s, tau)),
       y = expression("Z score " * z == hat(theta) / s)) +
  coord_cartesian(expand = FALSE) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    # unit(1, "null") lets the colorbar stretch to the full height of the panel area
    legend.key.height = unit(1, "null"),
    legend.key.width = unit(0.6, "cm")
  )
```

Lighter areas indicate where the weight of the data is higher, darker areas are where the weight of the data is lower.  The solid white curves are contours of $\lambda$ under the t prior, drawn at 0.1, 0.3, 0.5, 0.7, and 0.9.  The dotted white lines mark where a *normal* prior of the same standard deviation puts those same contours; because $\lambda = 1/(1 + r^2)$ takes no account of the point estimate, they are vertical, sitting at $r = \sqrt{1/\lambda - 1}$.  Each dotted line pairs with the solid contour at the same value, so the horizontal distance between the two is the difference the heavy tails make.  Notice that, conditional on $r = s / \tau$ remaining constant, the data get more weight as the z score $z$ gets larger.  This is exactly the kind of behavior we want from the prior to fix the prior data conflict.  Interestingly, there is adaptive shrinkage even when $\nu$ is large enough for the prior to be approximately normal (see the $\nu = 50$ subplot, bottom right).

Now, these are nice plots but I don't think they really convey what is happening.  Rather, if we plot z score on the $x$-axis and shrunk z score on the $y$-axis, then the behavior becomes really clear.  

```{r fig-shrunk-z}
#| fig-cap: |
#|   Shrunk z score $\lambda z$ against the observed z score. Rows vary the degrees of
#|   freedom, columns the ratio of uncertainty. The dotted line is a normal prior matched
#|   on standard deviation, and the grey diagonal is no shrinkage at all.
#| fig-width: 9
#| fig-height: 8.5

shrunk <- expand_grid(
  nu = c(3, 10, 20, 50),
  r = c(0.5, 2, 4),
  z = seq(-10, 10, length.out = 200)
) |>
  mutate(
    # With s = 1 the posterior mean is already on the z scale
    shrunk_z = pmap_dbl(list(z, r, nu), post_mean),
    normal_z = z / (1 + r^2)
  )

shrunk_long <- shrunk |>
  pivot_longer(c(shrunk_z, normal_z), names_to = "prior", values_to = "shrunk") |>
  mutate(prior = if_else(prior == "shrunk_z", "Student's t", "Normal"))

# One curve per panel, so r is carried by the facet rather than by colour
ggplot(shrunk_long, aes(z, shrunk, linetype = prior)) +
  geom_abline(slope = 1, intercept = 0, colour = "gray80", linewidth = 0.4) +
  geom_line(linewidth = 0.9) +
  facet_grid(
    nu ~ r,
    labeller = label_bquote(rows = nu == .(nu), cols = r == .(r))
  ) +
  scale_linetype_manual(
    values = c("Student's t" = "solid", "Normal" = "22"), name = NULL
  ) +
  labs(
    x = expression("Z score " * z == hat(theta) / s),
    y = expression("Shrunk z score " * lambda * z)
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

In my opinion, this is the clearest plot of the shrinkage behavior of the Student's t prior and of how the various parameters affect it.  As the estimate $\hat \theta$ becomes more certain ($r$ gets smaller), the shrinkage gets smaller for all estimates.  This makes sense, because as $r$ gets smaller the likelihood gets more weight.  The degrees of freedom, $\nu$, control how far along the $z$ axis the prior keeps shrinking the estimate before easing off.  Small values of $\nu$ mean the Student's t adaptively shrinks at smaller values of $z$, while larger $\nu$ shrink $z$ more and more like the normal prior would.


## Conclusion

OK, so what?  Does this mean we should never use a normal prior? I'm not sure that conclusion is justified, so I'll give you my own thoughts on when you should use each.

First, it makes sense to think about when long tails can arise.  I can't say with perfect certainty when a heavy tail makes sense, but if you're brand new to A/B testing and many teams are trying to move the same metric, then I think a Student's t prior makes a lot of sense.  Many levers are being pulled; some may work and some may _really_ work.  A heavier tail can allow teams to get credit where they deserve it, and shrink where shrinkage is needed.  Indeed, Azevedo et. al report in their 2020 paper [A/B Testing with Fat Tails](https://www.jstor.org/stable/27347381?casa_token=5KkdGjmNuQEAAAAA:4dcrPyh0AgX3RPUmKHDy1NGHdxsiAhdXhMxtdJnCFUheO_UGA3kzZtIbYbXXMjIbHHI0xp2kYllF7_IhrVGws8_XJpxbFjA-jyA6HRm5tFW8yMwnuS9vyg) that the distribution of observed effects for experiments run at Bing is fatter tailed than a normal may allow.  These results are marginal, meaning they are not broken out by team or surface.  If what we observe at Bing is true across the industry, then maybe a Student's t makes sense when we can't create priors for our particular team.

That being said, when a team has run many, many A/B tests and has a good corpus of experiments which can be thought of as exchangeable (because, for example, all experiments are run on the same surface by the same team), then a normal prior can make sense because very extreme effects are unlikely (for example, because the surface has been highly optimized and the low-hanging fruit has been plucked).  

The idea is not to say we should universally use one or the other, but rather to be transparent about our assumptions and why we are using one prior over another.